---
title: "MCMC Part 2"
author: "Julia Wrobel"
format:
  beamer:
    theme: metropolis
    aspectratio: 169
    fontsize: 14pt
    slide-level: 2
execute:
  echo: true
  eval: false
  warning: false
  message: false
header-includes:
  - \usepackage{xcolor}
  - \newcommand{\red}[1]{\textcolor{red}{#1}}
  - \newcommand{\green}[1]{\textcolor{green}{#1}}
---



## Overview

```{r, echo = FALSE, eval = TRUE}
library(tidyverse)
```


Today, we cover:


- Gibb's sampling
- Hamiltonian MC (briefly)


Resources:

- Peng Advanced Stat Computing Chapter 7
- [Book chapter on Hamiltonian MC](https://arxiv.org/pdf/1206.1901)

---




## Single component Metropolis-Hastings

\fontsize{11pt}{12pt}\selectfont
The standard Metropolis algorithm updates the entire parameter vector at once with a single step. 

- Hence, a $p$-dimensional parameter vector must have a $p$-dimensional proposal distribution $q$. 
- However, it is sometimes simpler to update individual components of the parameter vector one at a time, in an algorithm known as
single component Metropolis-Hastings (SCMH).

Let $x^t = (x_1^t, x_2^t, \ldots, x_p^t)$ be a $p$-dimensional vector of parameter values at iteration $t$.

---

## Single component M-H

\fontsize{11pt}{12pt}\selectfont
The SCMH algorithm updates each parameter in $x$ one at a time. All that is needed to update the $i$th component of $x^t$ is a proposal distribution $q(x_i^*|x_i^t, x^t_{-i})$.  SCMH updates the $i$th component via the following steps: 


(1) Sample $x_i^*\sim q(x_i^*|x_i^t, x^t_{-i})$ as a proposal for component $i$.
(2) Let 

$$\alpha(x_i^*|x^t) =  \min\left\{\frac{\pi(x_i^*|x_{-i}^t)q(x_i^t|x_i^*, x^t_{-i})}{\pi(x_i^t|x_{-i}^t)q(x_i^*|x_i^t, x^t_{-i})}, 1\right\}$$

(3) Accept $x_i^*$ with probability $\alpha(x_i^*|x^t)$

(4) Repeat steps 1-3 $p$ times until each component is updated

---

## Gibb's sampling

- Special case of single component M-H where you always accept the candidate
- For sampling from multivariate distributions (2 or more parameters)
  - Relies on ability to calculate full conditionals


Useful when:

- Sampling from $\pi(\theta_1, \theta_2, \theta_3)$ is hard
- Sampling from $\pi(\theta_1|\theta_2, \theta_3)$, $\pi(\theta_2|\theta_1, \theta_3)$ , $\pi(\theta_3|\theta_1, \theta_2)$  is easy



::: notes
We will revisit how this is a special case of M-H
:::

---

## Gibb's sampling algorithm

Steps for each iteration:

1. Sample $\theta_1^{n+1} \sim \pi(\theta_1|\theta_2^n, \theta_3^n)$
2. Sample $\theta_2^{n+1} \sim \pi(\theta_2|\theta_1^{n+1}, \theta_3^n)$
3. Sample $\theta_3^{n+1} \sim \pi(\theta_3|\theta_1^{n+1}, \theta_2^{n+1})$


\vspace{5mm}

Do this for the desired number of samples, discard burn-in, and voila! You now have a sample from the joint distribution $\pi(\theta_1,\theta_2,\theta_3)$.

---

## Example: Bivariate normal distribution

Suppose we want to simulate $x = (x_1, x_2)$ from a bivariate Normal distribution with mean $\mu = (\mu_1, \mu_2)$ and covariance
matrix

$$\Sigma = \begin{pmatrix}
\sigma_1^2 & \rho \\
\rho & \sigma_2^2
\end{pmatrix}$$

we can make use of Gibbs sampling to simulate a Markov chain
that will converge to a bivariate Normal. 


::: notes
From Peng, kind of a dumb example because you can sample from bivariate normal directly. But, useful because you can visualize it.
:::

---


## Bivariate normal distribution

Goal is to simulate from joint distribution $\pi(x_1, x_2 | \mu, \Sigma)$. Complete conditionals are:

- $\pi(x_1|x_2, \mu, \Sigma) = N(\mu_1+\rho\frac{\sigma_1}{\sigma_2}(x_2-\mu_2), \sigma_1^2(1-\rho))$
- $\pi(x_2|x_1, \mu, \Sigma) = N(\mu_2+\rho\frac{\sigma_2}{\sigma_1}(x_1-\mu_1), \sigma_2^2(1-\rho))$

<br>
So, we sample 

1. $x_1^{t+1} \sim N(\mu_1+\rho\frac{\sigma_1}{\sigma_2}(x^t_2-\mu_2), \sigma_1^2(1-\rho))$
2. $x_2^{t+1} \sim N(\mu_2+\rho\frac{\sigma_2}{\sigma_1}(x_1^{t+1}-\mu_1), \sigma_2^2(1-\rho))$


---

## Bivariate normal distribution

\fontsize{8pt}{9pt}\selectfont
```{r, eval = TRUE}
gibbs_bivariate = function(x0, n = 1000,
                           mu = c(1, 5), rho = 0.5, 
                           sigma1 = 1, sigma2 = 2){
  x1 = x0[1]
  x2 = x0[2]
  mu1 = mu[1]
  mu2 = mu[2]
  
  x = matrix(NA, nrow = n, ncol = 2)
  for(i in 1:n){
    x[i,] = c(x1,x2)
    
    x1 = rnorm(1, mean = mu1 + rho * sigma1/sigma2 * (x2-mu2), 
               sd = sigma1 * sqrt(1-rho))
    x2 = rnorm(1, mean = mu2 + rho * sigma2/sigma1 * (x1-mu1), 
               sd = sigma2 * sqrt(1-rho))
  }
  colnames(x) = c("x1", "x2")
  return(x)
}
```

---

## Bivariate normal distribution

\fontsize{11pt}{12pt}\selectfont
```{r, eval = TRUE}
set.seed(21433)
gibbs_x = gibbs_bivariate(c(0,0), n = 10000)

colMeans(gibbs_x[1:100,])
colMeans(gibbs_x)
```

---



## Bivariate normal distribution


```{r, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10}
df = as_tibble(gibbs_x) %>%
  mutate(iteration = row_number()) %>%
  filter(iteration >1000) 


set.seed(223)
rho <- 0.5
mu <- c(1, 5)
Sigma <- matrix(c(1, rho, rho, 4), 2, 2)

# Create a grid to evaluate the true density
x_seq <- seq(min(df$x1) - 1, max(df$x1) + 1, length.out = 100)
y_seq <- seq(min(df$x2) - 1, max(df$x2) + 1, length.out = 100)
grid <- expand.grid(x = x_seq, y = y_seq)

# Add true bivariate normal density values
grid$density <- mvtnorm::dmvnorm(grid, mean = mu, sigma = Sigma)

# Plot: real data points + true density contours and fill
ggplot(df, aes(x = x, y = y)) +
  geom_point(aes(x = x1, y = x2), alpha = 0.3, color = "blue") +  # real data
  geom_contour(data = grid, aes(z = density), color = "black") +  # true contours
  geom_raster(data = grid, aes(x = x, y = y, fill = density), alpha = 0.5, interpolate = TRUE) +
  scale_fill_viridis_c() +
  theme_minimal() 
```

::: notes
Through away the first 1000 observations, plotted true density on top
:::

---

## Bivariate normal distribution


```{r, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10}
as_tibble(gibbs_x) %>%
  mutate(iteration = row_number()) %>%
  filter(iteration <= 10) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(aes(x = x1, y = x2),  color = "blue", size = 2) +  # real data
  geom_path(aes(x = x1, y = x2), alpha = 0.3, color = "blue") +  # real data
  geom_contour(data = grid, aes(z = density), color = "black") +  # true contours
  theme_minimal() 

```


---


## Proof: Gibbs sampling as M-H

Gibbs is a special case of single component M-H that always accepts the proposal. Define the following:

- $\pi(\theta_1, \theta_2)$: joint distribution we want to sample from
- $q(\theta_1^*|\theta_1^t,\theta_2^t)$: proposal distribution for new value of $\theta_1$

- Acceptance ratio:

$$\alpha(\theta_1^*|\theta^t) =  \min\left\{\frac{\pi(\theta_1^*,\theta_2^t)q(\theta_1^t|\theta_1^*, \theta^t_2)}{\pi(\theta_1^t,\theta_2^t)q(\theta_1^*|\theta_1^t, \theta_2^t)}, 1\right\}$$

- Choose full conditional as proposal, i.e.:

$$q(\theta_1^*| \theta_1^t,\theta_2^t) = \pi(\theta^*_1|\theta_2^t)$$

---


## Proof: Gibbs sampling as M-H


::: notes
See gibbs_proof.png
:::

---


## Bayesian linear regression

\fontsize{10pt}{11pt}\selectfont
$$y = X\beta + \epsilon; \epsilon \sim N(0, \sigma^2_yI_n)$$

- Need to make (prior) distributional assumptions about $\beta$, $\sigma^2_y$
  - $\beta \sim N(0, \sigma^2_bI_p)$
  - $\sigma^2_y \sim IG(A, B)$


- Joint posterior of interest is 
 
$$\pi(\beta, \sigma^2_y|y, x, A, B, \sigma^2_b)$$

- Need full conditionals
  - $\pi(\beta|\sigma^2_y, y, x, A, B, \sigma^2_b)$
  - $\pi(\sigma^2_y|\beta, y, x, A, B, \sigma^2_b)$

---

## Full conditionals

\fontsize{11pt}{12pt}\selectfont
$$\beta \mid \sigma^2_y, y \sim \mathcal{N}(\beta_{\text{post}}, \Sigma_{\text{post}})$$

$$\Sigma_{\text{post}} = \left( \frac{1}{\sigma_y^2} X^\top X + \frac{1}{\sigma_\beta^2} I_p \right)^{-1}$$

$$\beta_{\text{post}} = \left( X^\top X + \frac{\sigma_y^2}{\sigma_\beta^2} I_p \right)^{-1} \left( X^\top y \right)$$

$$\sigma^2_y \mid \beta, y \sim IG\left(A + n/2, B + \frac{1}{2}||y-x\beta||^2\right)$$



---

## Drawbacks of Gibbs and Random walk MH

- Tend to get stuck in small regions of the posterior for a long time 
  - Especially for large number of parameters 
  - I.E. slow mixing for correlated parameters 
  - Potentially low effective sample size
- Requires closed-form full conditionals

- Random walk Metropolis-Hastings can be slow because it slowly explores the state space

::: notes
If you don't have closed form you can use metropolis-within gibbs
:::

---

## Variations on Gibb's sampling

- **Blocked Gibb's sampling**: update blocks (or groups) of parameters jointly rather than one at a time
  - Helps improve mixing when variables are highly correlated
  - We are already doing this for Bayesian linear regression
- **Collapsed Gibb's sampling**: integrate/marginalize out some variables analytically before sampling
  - Reduces dimensionality and can improve convergence
- **Metropolis-within-Gibbs**: Uses Metropolis-Hastings steps for full conditionals that are not easily sampled from.
  - Commonly used in Bayesian hierarchical models


---

## Hamiltonian Monte Carlo

> Though originating in physics, Hamiltonian dynamics can be applied to most problems with continuous state spaces by simply introducing fictitious “momentum” variables.


> HMC alternates updates for momentum variables with Metropolis updates in which a new state is proposed by computing a trajectory according to Hamiltonian dynamics. A state proposed in this way can be
distant from the current state but nevertheless have a high probability of acceptance.

::: notes
quote from hamiltonian MC book chapter
::: 

---

## Hamiltonian Monte Carlo

- The first step is to define a Hamiltonian function in
terms of the probability distribution we wish to sample from. 
- Parameters of interest are called "position" variables
- Also introduce auxiliary "momentum" variables

---

## Hamiltonian dynamics

\fontsize{11pt}{12pt}\selectfont
> Hamiltonian dynamics has a physical interpretation that can provide useful intuitions. In two dimensions, we can visualize the dynamics as that of a frictionless puck that slides over a surface of varying height The state of this system consists of the position of the puck, given
by a 2D vector $q$, and the momentum of the puck (its mass times its velocity), given by a 2D vector $p$. The potential energy, $U(q)$, of the puck is proportional to the height of the surface
at its current position, and its kinetic energy, $K(p)$, is equal to $|p|
^2/(2m)$, where $m$ is the
mass of the puck. On a level part of the surface, the puck moves at a constant velocity, equal
to $p/m$. If it encounters a rising slope, the puck’s momentum allows it to continue, with its
kinetic energy decreasing and its potential energy increasing, until the kinetic energy (and hence $p$) is zero, at which point it will slide back down (with kinetic energy increasing and
potential energy decreasing).

---

## Hamiltonian dynamics

$$H(q,p) = U(q) + K(p)$$

- U(q): potential energy
- K(p): kinetic energy
  - Usually defined as $p^TM^{-1}p/2$, where $M$ is a "mass matrix"
  
::: notes
mass matrix is often a scalar multiple of the identity
:::
---


## Hamiltonian dynamics

The partial derivatives of the Hamiltonian determine how $q$ and $p$ change over time $t$, according to Hamilton's equations:

$$\frac{dq_i}{dt} = \frac{\partial H}{\partial p_i}$$

$$\frac{dp_i}{dt} = -\frac{\partial H}{\partial q_i}$$

---

## Hamiltonian dynamics in statistics

- **position** corresponds to parameters of interest
- **potential energy U(q)** is the (joint) log probability density of interest 
- **kinetic energy K(p)** is usually defined as $p^TM^{-1}p/2$, where $M$ is a "mass matrix"
  - Corresponds to zero-mean Gaussian with covariance $M$

Can be used to construct MCMC updates.

::: notes
leapfrog method is used to approximate the differential equations and choose step size
:::
---



## Gibbs sampler for Bayesian regression

I started implementing this myself... for lab we will go the rest of the way there and compare it to the results of `stan_glm` from the Bayesian Data Analysis lecture.

- My partial Gibb's sampler is in the file `gibbs_blr.R` and the lab file is `lab_gibbs.Rmd`.
- Compare effective sample size for your Gibb's sampler and for Stan's HMC algorithm
- Compare diagnostic plots as well

---

## Resources

- [Book chapter on Hamiltonian MC](https://arxiv.org/pdf/1206.1901)
- [Stan manual](https://mc-stan.org/docs/2_19/reference-manual/convergence.html)
- [No U-Turn Sampler (NUTS)](https://www.jmlr.org/papers/volume15/hoffman14a/hoffman14a.pdf)
---

