---
title: "MCMC Part 1"
author: "Julia Wrobel"
format:
  beamer:
    theme: metropolis
    aspectratio: 169
    fontsize: 14pt
    slide-level: 2
execute:
  echo: true
  eval: false
  warning: false
  message: false
header-includes:
  - \usepackage{xcolor}
  - \newcommand{\red}[1]{\textcolor{red}{#1}}
  - \newcommand{\green}[1]{\textcolor{green}{#1}}
---



## Overview

\fontsize{10pt}{11pt}\selectfont
```{r, echo = FALSE, eval = TRUE}
library(tidyverse)
```


Today, we cover:

 
- Markov chains
- MCMC
- Metropolis-Hastings

Announcements:

- Next week spring break!
- No class Wednesday, 3/18 (ENAR)
- Resume Bayesian Computing on 3/19
- Look out for email from Keven Haynes about the cluster


Resources:

- Peng Advanced Stat Computing Chapter 7

---


## Joint posterior distribution

Our goal is to learn about the joint posterior distribution

$$p(\beta, \sigma^2_y|y, X, hyperparameters)$$

\vspace{5mm}
      
This can be hard!

- Often the joint posterior is analytically intractable
- For the conjugate priors we use, though, the full conditionals were "easy"
- If we can't write down the joint posterior, maybe we can sample from it


::: notes
This is the main motivation for MCMC
:::

---

## MCMC


Basic idea:  How do I get a string of values from the posterior when I cannot write the posterior down?

- Markov chain Monte Carlo (MCMC) methods let you sample from complicated [posterior] distributions
  - a collection of tools used to sample from a target distribution



- Once you have a sample from the posterior distribution you
can use summary statistics (mean, variance, quantiles) to
do posterior inference


First we will review Markov chains

::: notes
Gibbs sampler is a special case of M-H sampler
You can think about this analogously to bootstrap in some ways- once you have a sample from a distribution, you can learn about that distribution
:::

---

## Markov chains

\fontsize{10pt}{11pt}\selectfont
A Markov chain is a stochastic process that evolves over time by transitioning into different states. 

- The sequence of states is denoted by $X_i$
- Transition between states is random and follows the *Markov property*:

$$P(X_t | X_{t-1}, X_{t-2}, \ldots, X_0) = P(X_t|X_{t-1})$$


\vspace{10mm}

FYI: this notation is for discrete distributions

::: notes
Another word for random is probabilistic. Using discrete distribution here to get at the general idea, but usually we will be working with continuous distributions.
- transition matrix also sometimes called transition Kernel
:::

---

## Markov chains

- **Core concept**: The probability of transitioning to a state at time $t$ depends only on the most recent previous state

- *State space*: collection of states that a Markov chain can visit
- *Transition matrix*: collection of probabilities that we move from one state to another

::: notes
Another word for random is probabilistic. Using discrete distribution here to get at the general idea, but usually we will be working with continuous distributions.
- transition matrix also sometimes called transition Kernel
:::

---

## Markov chain example


::: notes
Draw picture here with three state example, and it's transition matrix
:::

---

##  Markov chain properties

\fontsize{11pt}{12pt}\selectfont
$$P = \begin{bmatrix}
\frac{1}{2} & \frac{1}{2} & 0 \\
\frac{1}{2} & 0 & \frac{1}{2} \\
0 & \frac{1}{2} & \frac{1}{2}
\end{bmatrix}$$

Each entry of $P$ is the probability of going from state $i$ to state $j$, i.e.

$$P(X_{n+1} = j |X_n = i) = P_{ij}$$

- Sum of weights from outgoing arrows = 1
- Starting with one state and sampling a string of subsequent states is called a **random walk**
- Goal is to find  probabilities of being in each state

::: notes
Here is the transition matrix from the previous example
:::

---


##  Markov chain properties

Suppose we start the chain in state 3, with probability 1, so the initial probability distribution over the three states is $\pi_0 = (0,0,1)$.

- What is the probability distribution over the states after one iteration?


\vspace{5mm}

- What is the probability distribution over the states after $n$ iterations?

\vspace{10mm}

$\pi_*$ is called the stationary distribution. This is the vector of probabilities of being in each state that the chain converges to.


::: notes
- $\pi_1 = \pi_0P = (0, 1/2, 1/2)$
- $\pi_n = \pi_0PPP\times \ldots\times P = \pi_0  P^n$
- As $n$ increases, we expect this to settle into a steady state
:::

---

##
##  Markov chain limit theorem

\fontsize{11pt}{12pt}\selectfont
Basic limit theorem for Markov chains says 

$$||\pi_*-\pi_n|| \to 0 \mbox{ as } n\to \infty$$

where $||\cdot||$ is the total variation distance between two densities.

- another way of saying this is $\lim_{n\to\infty} \pi^n = \pi_*$ regardless of the starting state

In the long run, the chain "forgets" where it started and settles into a stable distribution over the states. That stable/stationary distribution is $\pi_*$, and the process spends a proportion $\pi_j$ of time in state $j$.

::: notes
This limit theorem holds under some necessary assumptions
:::

---

##  Markov chain assumptions

(1) The stationary distribution $\pi_*$ exists.

(2) The chain is **irreducible**: it is possible to reach any state from any other state (eventually)


\vspace{15mm}

(3) The chain is **aperiodic**: the chain does not cycle in a fixed period. A chain is aperiodic if it does not make deterministic visits to a subset of the state space.


::: notes
You don't have to be able to reach other states in just one step.
- periodicity means you can deterministically flip between states. Use pizza on Mondays example
If these assumptions hold, then Markov chain limit theorem holds and the stationary distribution is unique.
:::

---


## Time-reversible Markov chains

\fontsize{11pt}{12pt}\selectfont
In a reversible Markov chain, the process looks the same if you run it forward or backward in time.

- At equilibrium (the chain’s long-run steady state), the traffic between any two states is perfectly balanced. For any two states A and B, the probability of transitioning from A to B times the long-run probability of being in A equals the probability of going from B to A times the long-run probability of being in B


::: notes
Why do we care about this for MCMC?
:::

---



## Why Reversibility Matters for MCMC

You want samples from $p(\theta|y) = $ complicated posterior distribution. Sampling directly is hard. Instead:

1. Build a Markov chain
2. Ensure stationary distribution = target distribution
3. Run the chain for a while.
4. Visited states become samples from the target.


MCMC methods use Markov chains to sample from a target probability distribution. The goal is to construct a chain whose stationary distribution is the target distribution we want to sample. Reversibility ensures we have the correct stationary distribution.


---

  

## Why Reversibility Matters for MCMC

- Ensures correctness: target distribution is the stationary distribution
- Prevents systematic bias: no net directional drift in sampling
- Simplifies algorithm design and analysis (e.g., Metropolis-Hastings)
- Makes it easier to combine steps while preserving invariance
- Widely used in MCMC due to theoretical guarantees and reliability

---



## Review: rejection sampling



::: notes
What did we learn about this at a high level? What is the point of this?
- Just draw a picture here
- Try to provide visual intuition behind this, and why it is inefficient when M is large. Also, because samples are independent we don't learn anything from previous samples about how good they are
:::

---


## MCMC intuition

- Next sample depends on the last sample
- Samples are no longer independent
- Eventually Markov chain will simulate draws from target distribution
- If a Markov chain reaches a stationary distribution, it will stay there
  - Goal is to design a Markov chain whose stationary distribution is the target distribution

::: notes
MCMC is a general idea, not an algorithm. 
- draw picture for this too and connect it to rejection sampling
:::

---

## MCMC

\fontsize{11pt}{12pt}\selectfont
Term MCMC encompasses broad array of techniques that have in common a few key ideas:

1. We want to sample form a complicated density or probability mass function $\pi$, often a posterior distribution. Assumption is we can evaluate $\pi$ but not *directly* sample from it
2. We know that certain stochastic processes called *Markov chains* will converge to a stationary distribution.
  - Simulating from a Markov chain for long enough will eventually gives us a sample from the chain's stationary distribution

3. Given the functional form of the density $\pi$, we want to construct a Markov chain that has $\pi$ as its stationary distribution
4. We want to sample values from the MC such that the sequence of values $\{x_n\}$ generated by the chain converges in distribution to $\pi$.

::: notes
If we can construct a Markov chain whose stationary distribution is our target $\pi$ then we can sample from $\pi$.
Markov chains converge to their stationary distribution if it exists and specific conditions are satisfied
- How you design your markov chain differs from method to method
:::

---



## Metropolis-Hastings algorithm

\fontsize{10pt}{11pt}\selectfont
Suppose we want to take a sample $x \sim \pi(x)$ using MCMC. 

- Metropolis-Hastings is a popular algorithm (circa 1970) for doing this. 


Some terminology:

- **target distribution** is $\pi(x)$, the distribution I want to sample from
- **proposal distribution** is $q(y|x)$, some probability distribution that is easy to sample from that has the same support as $\pi$
  - You can equivalently think of this as a transition probability to get from $x$ to the next state $y$
- **acceptance probability** aka **acceptance ratio**, denoted $\alpha(y|x)$, is the probability of transitioning to a new stage in the Markov chain



::: notes
- note that these things were invented a long time a go but it took quite a while for them to take off, until computers got much faster. Hastings didn't even get tenure!
:::

---

## Steps of Metropolis-Hastings 

\fontsize{11pt}{12pt}\selectfont
The Metropolis-Hastings procedure is an iterative algorithm where at each iteration, there are three steps. Suppose we are currently in state $x$ and want to know how to move to the next state in the state space (which we will call $y$).

(1) Simulate a *candidate* value $y\sim q(y|x)$ from the proposal distribution, conditional on current state $x$.

(2) Calculate the acceptance ratio, 

$$\alpha(y|x) = \min\left\{\frac{\pi(y)q(x|y)}{\pi(x)q(y|x)}, 1\right\}$$
(3) Simulate $u\sim Unif(0,1)$. If $u\le \alpha(y|x)$, then the next state is equal to $y$. Otherwise, the next state is still $x$.

::: notes
Recall our hope is that our Markov chain will, after many simulations, converge to the stationary distribution.
:::

---

## Simple example

Say we want to sample from the probability distribution $p(x)$ given below:

\begin{align}
p(x) &= \frac{\exp(-x^2)(2 + \sin(5x) + \sin(2x))}{\int_{-\infty}^{\infty}\exp(-x^2)(2 + \sin(5x) + \sin(2x))dx}\\[3mm]
&\propto \exp(-x^2)(2 + \sin(5x) + \sin(2x))
\end{align}

::: notes
- bit of a contrived example to show how it works but we will do a more complex example next
Usually problems like this we just call the denominator the normalizing constant
:::

---

## Simple example

```{r, echo = FALSE,  eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10, warning = FALSE}
p <- function(x) {
  exp(-x^2) * (2 + sin(5 * x) + sin(2 * x))
}

# Compute normalization constant Z ≈ ∫ p(x) dx over [-Inf, Inf]
Z <- integrate(p, lower = -Inf, upper = Inf)$value

# Normalized density
f <- function(x) {
  p(x) / Z
}


x_vals <- seq(-3, 3, length.out = 1000)
f_vals <- f(x_vals)
df <- data.frame(x = x_vals, f = f_vals)

ggplot(df, aes(x = x, y = f)) +
  geom_line(color = "blue", size = 1) +
  labs(title = expression(p(x) == frac(e^{-x^2}*(2 + sin(5*x) + sin(2*x)), Z)),
       y = expression(p(x)), x = "x") +
  theme_minimal()
```


---

## Simple example

\fontsize{11pt}{12pt}\selectfont
Domain of $x$ is unrestricted.  Let $x$ be our current value and $y$ the next value in the chain.

- To generate $y$  we will add a $N(0, 1)$ random variable to our current value of $x$
  - This implies the proposal distribution $q(y|x)$ follows $y \sim N(x, 1)$
  
\vspace{5mm}  

The more general idea here is to let $y = x + \epsilon$, where $\epsilon \sim g$ and $g$ is a probability density symmetric about 0.

- Sometimes called a "random walk" Metropolis-Hastings
- Using a symmetric proposal distribution is also called the Metropolis algorithm


::: notes
Notice that the distribution we sample from depends on the current state of the Markov chain
:::

---

## Acceptance ratio step


$$\alpha(y|x) = \min\left\{\frac{\pi(y)q(x|y)}{\pi(x)q(y|x)}, 1\right\}$$


\vspace{20mm}

::: notes
Note here why the normalizing constant will go away, and why the $\alpha$ simplifies
:::

---

## Acceptance ratio step

$$\alpha(y|x) = \min\left\{\frac{\pi(y)}{\pi(x)}, 1\right\}$$
First, we calculate this ratio. 

\vspace{5mm}

Then simulate $u\sim Unif(0,1)$. 

- If $u\le \alpha(y|x)$, then the next state is equal to $y$. Otherwise, the next state is still $x$.

---

## Simple example

```{r, echo = FALSE,  eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10, warning = FALSE}
p <- function(x) {
  exp(-x^2) * (2 + sin(5 * x) + sin(2 * x))
}

# Compute normalization constant Z ≈ ∫ p(x) dx over [-Inf, Inf]
Z <- integrate(p, lower = -Inf, upper = Inf)$value

# Normalized density
f <- function(x) {
  p(x) / Z
}


x_vals <- seq(-3, 3, length.out = 1000)
f_vals <- f(x_vals)
df <- data.frame(x = x_vals, f = f_vals)

ggplot(df, aes(x = x, y = f)) +
  geom_line(color = "blue", size = 1) +
  labs(title = expression(p(x) == frac(e^{-x^2}*(2 + sin(5*x) + sin(2*x)), Z)),
       y = expression(p(x)), x = "x") +
  theme_minimal()
```


---
  
## Simple example implementation

\fontsize{11pt}{12pt}\selectfont
```{r, eval = TRUE}
target = function(x){
  exp(-x^2) * (2 + sin(5 * x) + sin(2 * x))
}

metropolis_step = function(x, sigma) {
  y = rnorm(1, mean = x, sd = sigma)
  accept_prob <- min(1, target(y) / target(x))
  u = runif(1)
  if(u <= accept_prob) {
    value = y
    accepted = TRUE
  } else {
    value = x
    accepted = FALSE
  }
  
  tibble(value = value, accepted = accepted)
  
}
```

---

## Simple example implementation

\fontsize{11pt}{12pt}\selectfont    
```{r, eval = TRUE}
metropolis_sampler = function(initial_value, 
                               n = 1000, 
                               sigma = 1) {
  
  results = as.list(rep(NA, n))
  current_state = initial_value
  for(i in 1:n) {
    out = metropolis_step(current_state, sigma)
    current_state = out$value
    results[[i]] = out
  }
  results = do.call(rbind, results)
  results
}
```

---

## Simple example implementation

\fontsize{11pt}{12pt}\selectfont   
```{r, eval = TRUE}
set.seed(34634634)
chain1 = metropolis_sampler(initial_value = 0, n = 1000)

head(chain1)
```

::: notes
Could be helpful to plot rolling acceptance ratio 
:::

---


## Visualizing results

\fontsize{11pt}{12pt}\selectfont   
```{r, eval = TRUE}
set.seed(3463)
chain2 = metropolis_sampler(initial_value = 0, n = 1000)
chain3 = metropolis_sampler(initial_value = 10, n = 1000)
chain4 = metropolis_sampler(initial_value = -20, n = 1000)
```


---

## Visualizing results


```{r, echo = FALSE,  eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10}
results1 = bind_rows(mutate(chain1, chain = 1),
          mutate(chain2, chain = 2),
          mutate(chain3, chain = 3),
          mutate(chain4, chain = 4)) 


results1 %>%
  filter(accepted == TRUE) %>%
  ggplot(aes(value)) +
  geom_histogram() +
  facet_wrap(~chain, scales = "free")
```


---

## Acceptance probabilities

```{r, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10, warning = FALSE}
results1 %>%
  group_by(chain) %>%
  mutate(accepted = as.numeric(accepted),
         prob = zoo::rollmean(accepted, k = 20, fill = NA),
         observation = row_number()) %>%
  ungroup() %>%
  ggplot(aes(observation, prob)) +
  geom_line() +
  facet_wrap(~chain)
```

---

## Some terminology

- **Burn-in**: the initial period of an MCMC simulation discarded to allow the chain to reach its stationary distribution.
- **Lag**: refers to the number of steps between samples used to reduce autocorrelation and improve the independence of the retained draws.
- **Mixing rate**:  refers to how quickly an MCMC chain explores the target distribution, with better mixing indicating faster decorrelation between samples.
 - **Proposal width** is a parameter of the proposal you can vary to tune this


::: notes
We will look at how each of these affect the sampler
:::

---


## Burn-in

Convergence occurs regardless of our starting point (in theory), so we can usually pick any reasonable values in the parameter space for initial value(s).

- The farther away you are from high density regions the longer it will take to reach the stationary distribution (i.e. target distribution $\pi(x)$)
- Samples      you generate early on will probably not follow $\pi(x)$
- Generally we throw out a certain number of the first draws, known as **burn-in** to make sure our sample is closer to the stationary distribution and less dependent on starting values


::: notes
What is a good starting point for our distribution example?
- Not always obvious how much burn in to discard, and the decision is fairly arbitrary
- plots can help
:::

---


## Burn-in: $x_0 = -20$

```{r, cache = TRUE, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 9, message = FALSE}
set.seed(23423)
chain5 = metropolis_sampler(initial_value = -20, n = 10000)
chain5 = chain5 %>%
  filter(accepted == TRUE) %>%
  mutate(observation = row_number(),
         quintile = case_when(observation < 2000 ~ 1,
                              observation < 4000 ~ 2,
                              observation < 6000 ~ 3,
                              observation < 8000 ~ 4,
                              TRUE ~ 5),
         quintile = factor(quintile, levels = 1:5, labels = c("<2000", "2000-4000", 
                                                              "4000-6000", "6000-8000", "8000-1000")))

chain5 %>%
  ggplot(aes(value)) +
  geom_histogram() +
  facet_wrap(~quintile, scales = "free", ncol = 1)
```

---

## Lag

\fontsize{11pt}{12pt}\selectfont   
This is essentially the number of iterations we run the sampler for between successive samples.

- MCMC samples are not independent! By definition, each sample depends on the previous sample
  - This means there is a time-dependent autocorrelation
- One way to reduce correlation in your samples and have draws that are more independent is to allow several iterations of the sampler to elapse in between successive samples. 
- This is the lag
- More of a problem when the MCMC cannot make big jumps within the parameter space


::: notes
When can this last point occur?  Small variance of your proposal distribution
:::

---


## Lag: autocorrelation plots

```{r, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10}
acf_vals <- acf(chain5$value, plot = TRUE)

```


---

## Mixing rate

In practice the choice of proposal distribution matters a lot. 

- If it’s too narrow, your sampler will have an extremely high acceptance rate on average, but it will move around extremely slowly, that is, 
  - i.e. it has a low *mixing rate*. 
- If it’s too wide, the acceptance rate becomes too low and the chain gets stuck on specific values for long periods of time. 

- Varying the proposal width can affect this

---


## Effect of proposal width

\fontsize{11pt}{12pt}\selectfont   
```{r, eval = TRUE}
set.seed(34634)
chain_1 = metropolis_sampler(initial_value = 0, n = 1000, sigma = 1)
chain_.025 = metropolis_sampler(initial_value = 0, n = 1000, sigma = .025)
chain_50 = metropolis_sampler(initial_value = 0, n = 1000, sigma = 50)

```

---


## Effect of proposal width



```{r, echo = FALSE,  eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10,message = FALSE}
results = bind_rows(mutate(chain_1, sigma = 1),
          mutate(chain_.025, sigma = .025),
          mutate(chain_50, sigma = 50)) %>%
  group_by(sigma) %>%
  mutate(iteration = row_number()) %>%
  ungroup() %>%
  mutate(sigma = factor(sigma))


results %>%
  filter(accepted == TRUE) %>%
  ggplot(aes(value)) +
  geom_histogram() +
  facet_wrap(~sigma, labeller = label_both)
```

---

## Effect of proposal width

```{r, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10, warning = FALSE}
results %>%
  group_by(sigma) %>%
  mutate(accepted = as.numeric(accepted),
         prob = zoo::rollmean(accepted, k = 20, fill = NA),
         observation = row_number()) %>%
  ungroup() %>%
  ggplot(aes(observation, prob)) +
    ggtitle("acceptance probabilities") +
  geom_line(aes(color = sigma))
```

---

## Convergence

We will not cover the convergence theory behind Metropolis chains in
detail, but ...

- The Markov process generated under this procedure is ergodic
(irreducible and aperiodic) and has a unique limiting distribution
(stationary distribution).
- E rgodicity means that the chain can move anywhere at each
step, which is ensured if $q(y|x)>0$ everywhere

- By construction, Metropolis chains are reversible, so that  $\pi(y|x)$ is the stationary distribution

---

## Convergence diagnostics

\fontsize{11pt}{12pt}\selectfont   
Convergence diagnostics are used to assess if your chain has converged.

- Cannot guarantee if a chain has converged, but they can indicate that it has not converged
- Not specific to Metropolis-Hastings, apply to MCMC more generally

\vspace{5mm}

- Trace plot
- Autocorrelation plot
- Density plots for multiple chains
- Effective sample size

---

## Trace plot

Shows sample values of an MCMC parameter over iterations

- Plot the whole chain (both accepted and rejected samples)

\vspace{5mm}

The chain should: 

- fluctuate around a stable mean without visible trends or drifts over time
- Avoid long streaks of similar values (these indicate high autocorrelation or poor mixing)
- Explore the full parameter space


---

## Trace plot

```{r, echo = FALSE,  eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10, warning = FALSE}
results %>%
  ggplot(aes(iteration, value)) +
  geom_line() +
  facet_wrap(~sigma, labeller = label_both)
```


---

## Autocorrelation plot

Shows correlation between samples at different lags. Lower autocorrelation means better mixing and more efficient sampling.

```{r, echo = FALSE, eval= TRUE,  fig.align='center', fig.height = 4, fig.width = 10, warning = FALSE}
acf_df <- function(chain, name) {
  acf_res <- acf(chain, plot = FALSE)
  tibble(
    lag = acf_res$lag[, 1, 1],
    acf = acf_res$acf[, 1, 1],
    sigma = name
  )
}

acfs = bind_rows(acf_df(chain_1$value, "1"), 
          acf_df(chain_.025$value, ".025"), 
          acf_df(chain_50$value, "50"))

acfs %>%
  ggplot(aes(x = lag, y = acf)) +
  geom_segment(aes(xend = lag, yend = 0), color = "steelblue") +
  geom_hline(yintercept = 0) +
  facet_wrap(~ sigma, nrow = 1, labeller = label_both) +
  theme_minimal()
```


 
---


## Density plots



```{r, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10, warning = FALSE}
results %>%
  filter(accepted == TRUE) %>%
  ggplot(aes(value)) +
  geom_density() +
  facet_wrap(~sigma, labeller = label_both)
```


::: notes
For these just plot accepted samples.
:::

---


## Density plots


```{r, echo = FALSE, eval = TRUE, fig.align='center', fig.height = 5, fig.width = 10, warning = FALSE}
results1 %>%
  filter(accepted == TRUE) %>%
  ggplot(aes(value)) +
  xlim(-5,5) +
  geom_density() +
  facet_wrap(~chain, labeller = label_both)
```


---


## Effective sample size

\fontsize{11pt}{12pt}\selectfont 
The effective sample size translates the number of MCMC samples
$S$ into an equivalent number of independent samples.

$$ESS = \frac{S}{1 + 2\sum_k\rho_k}$$

where $\rho_k$ is the lag $k$ autocorrelation.

- Can calculate ESS using `coda::effectiveSize`.

---

## Effective sample size




```{r, eval = TRUE}
library(coda)

effectiveSize(chain_.025$value) %>% as.numeric()
effectiveSize(chain_1$value) %>% as.numeric()
effectiveSize(chain_50$value) %>% as.numeric()
```


---



## General diagnostic rules

- Run multiple chains with different starting values
- Discard the first 1000 or so samples (depending on the problem)
- Check diagnostic plots!

- There are other diagnostic statistics (Gelman-Rubin, Geweke) that we haven't discussed

---




## Resources

- [Cute Markov chain example](https://www.youtube.com/watch?v=i3AkTO9HLXo&ab_channel=NormalizedNerd)
- [Intro MCMC video](https://www.youtube.com/watch?v=yApmR-c_hKU&ab_channel=ritvikmath)


