---
title: "Lab: Building an R Package"
subtitle: "BIOS 731: Advanced Statistical Computing"
author: "Julia Wrobel"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
hitheme: tomorrow
highlighter: highlight.js
---

```{r setup, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
```

# Overview

In this lab you will build a small R package called `projr` — a minimal project scaffolding tool inspired by [`workflowr`](https://workflowr.github.io/workflowr/). You start with three working R scripts and progressively turn them into a proper package with documentation and tests.

The three functions you will package are:

- `proj_start(name, path)` — creates a standardized project directory
- `proj_new_analysis(name, project_path)` — creates a new `.Rmd` file in `analysis/`
- `proj_status(project_path)` — prints a summary of files in the project

---

# Part 0: Try the scripts first

Source the three scripts from `labs/projr_scripts/` and try the functions:

```{r}
source("projr_scripts/proj_start.R")
source("projr_scripts/proj_new_analysis.R")
source("projr_scripts/proj_status.R")

proj_start("my_first_project", path = tempdir())
proj_new_analysis("exploratory", project_path = file.path(tempdir(), "my_first_project"))
proj_status(project_path = file.path(tempdir(), "my_first_project"))
```

Confirm you understand what each function does before moving on.

---

# Part 1: Create the package skeleton

Use `usethis` to create the package. Run this **outside** of any existing R project:

```{r}
library(usethis)
library(devtools)

create_package("~/projr")
```

This opens a new RStudio session inside the `projr/` project. All remaining steps happen there.

Set up a license and git:

```{r}
use_mit_license()
use_git()
```

---

# Part 2: Add functions

Create one `.R` file per function:

```{r}
use_r("proj_start")
use_r("proj_new_analysis")
use_r("proj_status")
```

Copy the body of each function from the starter scripts into the corresponding file. Then load everything and test interactively:

```{r}
load_all()

proj_start("test_proj", path = tempdir())
proj_status(file.path(tempdir(), "test_proj"))
```

---

# Part 3: Document with roxygen2

Add roxygen2 documentation above each function. Here is a template for `proj_start`:

```{r}
#' Start a new project
#'
#' Creates a standardized directory structure for a new research project,
#' including subdirectories for data, R scripts, analysis, results, and docs.
#'
#' @param name Name of the project (also used as the directory name).
#' @param path Parent directory where the project folder will be created.
#'   Defaults to the current working directory.
#'
#' @return The path to the new project directory (invisibly).
#' @export
#'
#' @examples
#' proj_start("my_project", path = tempdir())
proj_start <- function(name, path = ".") { ... }
```

Write similar documentation for `proj_new_analysis` and `proj_status`, then generate the help files:

```{r}
document()
```

Check that `?proj_start` works and that `NAMESPACE` now contains `export(proj_start)`.

---

# Part 4: Add a dependency

`proj_status` uses `sprintf()` and `cat()` from base R — no imports needed. But now extend `proj_new_analysis` to open the file in RStudio after creating it, using `rstudioapi`:

```{r}
use_package("rstudioapi", "Suggests")
```

Add the following lines at the end of `proj_new_analysis()`, after `writeLines()`:

```{r}
if (rstudioapi::isAvailable()) {
  rstudioapi::navigateToFile(filename)
}
```

Because `rstudioapi` is in `Suggests` (not `Imports`), guard the call with `rstudioapi::isAvailable()` so the function still works outside RStudio. Re-run `load_all()` and test.

---

# Part 5: Write tests

Set up `testthat`:

```{r}
use_testthat()
use_test("proj_start")
use_test("proj_status")
```

Fill in `tests/testthat/test-proj_start.R`:

```{r}
test_that("proj_start creates expected directories", {
  tmp <- tempdir()
  proj_start("testproj", path = tmp)

  project_path <- file.path(tmp, "testproj")
  expect_true(dir.exists(project_path))

  for (folder in c("data", "R", "analysis", "results", "docs")) {
    expect_true(dir.exists(file.path(project_path, folder)))
  }

  expect_true(file.exists(file.path(project_path, "README.md")))
})

test_that("proj_start errors if directory already exists", {
  tmp <- tempdir()
  proj_start("testproj2", path = tmp)
  expect_error(proj_start("testproj2", path = tmp))
})
```

Write at least one test for `proj_status` in `test-proj_status.R`. Then run:

```{r}
test()
```

---

# Part 6: Check the package

```{r}
check()
```

Work through any warnings or notes until you reach 0 errors, 0 warnings, 0 notes.

Common issues to watch for:

- Missing `@export` on a function you intended to be user-facing
- An example that calls `proj_start()` without writing to `tempdir()` (will fail because it tries to write during check)
- Undocumented arguments

---

# Stretch: Add a fourth function

Add `proj_build(project_path)` that renders all `.Rmd` files in `analysis/` to HTML using `rmarkdown::render()`. You will need to:

1. `use_r("proj_build")`
2. `use_package("rmarkdown")`
3. Write the function, document it, and export it
4. Write at least one test
5. Re-run `check()`
