---
title: "Rcpp"
author: "Julia Wrobel"
format:
  beamer:
    theme: metropolis
    aspectratio: 169
    fontsize: 14pt
    slide-level: 2
execute:
  echo: true
  eval: false
  warning: false
  message: false
header-includes:
  - \usepackage{xcolor}
  - \newcommand{\red}[1]{\textcolor{red}{#1}}
  - \newcommand{\green}[1]{\textcolor{green}{#1}}
---

```{r, echo = FALSE, eval = TRUE}
library(tidyverse)
library(microbenchmark)
```

## Overview

Today, we cover:

- Finish up R packages 
- RCPP


Announcements

- HW4 due April 15, 10:00AM
- No class tomorrow, April 9
- Email me to set up office hours



---


## About Rcpp

>"Sometimes R code just isn’t fast enough. You’ve used profiling to figure out where your bottlenecks are, and you’ve done everything you can in R, but your code still isn’t fast enough."  - Hadley Wickham



- In this lecture you’ll learn how to improve performance by rewriting key functions in C++
- Comes by way of the [Rcpp package](https://www.rcpp.org/)
  - Makes it simple to connect C++ code to R

---

## About Rcpp

\fontsize{10pt}{11pt}\selectfont
Typical bottlenecks that C++ can address include:


- Recursive functions, or problems which involve calling functions millions of times. The overhead of calling a function in C++ is much lower than that in R.
- Problems that require advanced data structures and algorithms that R doesn’t provide. Through the standard template library (STL), C++ has efficient implementations of many important data structures, from ordered maps to double-ended queues.
- Loops that can’t be easily vectorised because subsequent iterations depend on previous ones.
- Linear algebra stuff

::: notes
We're just going to talk a bit about some of the simpler stuff you can do with Rcpp
:::

---

## A note on Rcpp

You don't really need to know C++ to use Rcpp.

- But... Rcpp is notoriously annoying to debug.


::: notes
It took me over an hour to get it working on this machine since i haven't used it on the machine as a super simple example, even though I have used it previously. Actually it was 35 minutes but it felt like forever.
:::

---

## Getting started with C++

\fontsize{10pt}{11pt}\selectfont
`cppFunction()` allows you to write C++ functions in R:

```{r, eval = TRUE}
library(Rcpp)

cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')

# add works like a regular R function
add(1, 2, 3)

```

::: notes
When you run this code, Rcpp will compile the C++ code and construct an R function that connects to the compiled C++ function. We’re going to use this simple interface to learn how to write C++.
- Talk about whats going on here
:::

---


## Basic syntax

Translate R functions to their C++ equivalents. Gonna go through some increasingly complex examples:

- No inputs and a scalar output
- Scalar input and scalar output
- Vector input and scalar output
- Vector input and vector output
- Matrix input and vector output

---


## No inputs, scalar output

\fontsize{10pt}{11pt}\selectfont
This function has no arguments and always returns the integer 1:

```{r}
one <- function(){1L}
```

The equivalent C++ function is:

```{r, eval = FALSE}
int one() {
  return 1;
}
```

We can compile and use this from R with `cppFunction`:

```{r, eval = FALSE}
cppFunction('int one() {
  return 1;
}')
```


---


## No inputs, scalar output

\fontsize{10pt}{11pt}\selectfont
Reveals important differences between C++ and R:

- The syntax to create a function looks like the syntax to call a function; you don’t use assignment to create functions as you do in R.
- You must declare the type of output the function returns (`int` for scalar integer).
- Scalars and vectors are different
  - Common types of vectors are: `NumericVector`, `IntegerVector`, `CharacterVector`, and `LogicalVector`  
  - The scalar equivalents of numeric, integer, character, and logical vectors are `double`, `int`, `String`, and `bool`
- You must use an explicit `return` statement to return a value from a function.
- Every statement is terminated by a ;

---

## Scalar input, scalar output


\fontsize{8pt}{9pt}\selectfont
The next example function implements a scalar version of the `sign()` function which returns 1 if the input is positive, and -1 if it’s negative:

```{r, eval = FALSE}
signR <- function(x) {
  if (x > 0) {
    1
  } else if (x == 0) {
    0
  } else {
    -1
  }
}

cppFunction('int signC(int x) {
  if (x > 0) {
    return 1;
  } else if (x == 0) {
    return 0;
  } else {
    return -1;
  }
}')
```


::: notes
- We declare the type of each input in the same way we declare the type of the output. While this makes the code a little more verbose, it also makes it very obvious what type of input the function needs.
:::

---


## Vector input, scalar output


\fontsize{9pt}{10pt}\selectfont
One big difference between R and C++ is that the cost of loops is much lower in C++. In R we would never use the following code. But in C++, loops have very little overhead, so it’s fine to use them. 

```{r, eval = TRUE}
sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}
```

```{r, eval = TRUE}
cppFunction('double sumC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}')
```



::: notes
The following R code is terrible because we would never do this in a loop, we would use vectorization. However, this shows you how it works in both.
- To find the length of the vector, we use the .size() method, which returns an integer. C++ methods are called with . 
:::

---

## Vector input, scalar output


\fontsize{10pt}{11pt}\selectfont

```{r, eval = TRUE}
x <- runif(1e3)
microbenchmark(
  sum(x),
  sumC(x),
  sumR(x)
)
```


---

## Vector input, vector output


\fontsize{10pt}{11pt}\selectfont
Next we’ll create a function that computes the Euclidean distance between a value and a vector of values:

```{r}
pdistR <- function(x, ys) {
  sqrt((x - ys) ^ 2)
}
```

```{r}
cppFunction('NumericVector pdistC(double x, NumericVector ys) {
  int n = ys.size();
  NumericVector out(n);

  for(int i = 0; i < n; ++i) {
    out[i] = sqrt(pow(ys[i] - x, 2.0));
  }
  return out;
}')
```

::: notes
- We create a new numeric vector of length n with a constructor: NumericVector out(n).
:::

---


## Matrix input, vector output


\fontsize{10pt}{11pt}\selectfont
Each vector type has a matrix equivalent: `NumericMatrix`, `IntegerMatrix`, `CharacterMatrix`, and `LogicalMatrix`. For example, we could create a function that reproduces `rowSums()`:

```{r, eval = TRUE}
cppFunction('NumericVector rowSumsC(NumericMatrix x) {
  int nrow = x.nrow(), ncol = x.ncol();
  NumericVector out(nrow);

  for (int i = 0; i < nrow; i++) {
    double total = 0;
    for (int j = 0; j < ncol; j++) {
      total += x(i, j);
    }
    out[i] = total;
  }
  return out;
}')
set.seed(1014)
x <- matrix(sample(100), 10)
rowSumsC(x)
```


::: notes
- In C++, you subset a matrix with (), not [].
- Use .nrow() and .ncol() methods to get the dimensions of a matrix.
:::

---

## Using sourceCpp

\fontsize{10pt}{11pt}\selectfont
So far, we’ve used inline C++ with `cppFunction()`. For real problems, it’s usually easier to use stand-alone C++ files and then source them into R using `sourceCpp()`.


Your stand-alone C++ file should have extension `.cpp` and  start with:

```{cpp}
#include <Rcpp.h>
using namespace Rcpp;
```

And for each function that you want available within R, you need to prefix it with:

```{cpp}
// [[Rcpp::export]]
```

Note that the space is mandatory.

---

## Using sourceCpp

To compile the C++ code, use `sourceCpp("path/to/file.cpp")`.

- This will create the matching R functions and add them to your current session.


You can embed R code in special C++ comment blocks. This is really convenient if you want to run some test code:

```{cpp}
/*** R
# This is R code
*/
```

---


## Using sourceCpp


\fontsize{8pt}{9pt}\selectfont
```{cpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}

/*** R
library(microbenchmark)
x <- runif(1e5)
microbenchmark(
  mean(x),
  meanC(x)
)
*/
```


::: notes
Save and run on the fly
:::

---


## Rcpp sugar


\fontsize{10pt}{11pt}\selectfont
Rcpp provides a lot of syntactic “sugar” to ensure that C++ functions work very similarly to their R equivalents. In fact, Rcpp sugar makes it possible to write efficient C++ code that looks almost identical to its R equivalent. If there’s a sugar version of the function you’re interested in, you should use it: it’ll be both expressive and well tested.


Sugar functions can be roughly broken down into

- arithmetic and logical operators
- logical summary functions
- vector views
- other useful functions

---

## Rcpp sugar


\fontsize{10pt}{11pt}\selectfont
All the basic arithmetic and logical operators are vectorised: : +, *, -, /, pow, <, <=, >, >=, ==. For example a simpler version of our `pdistC()` function:

```{r}
pdistR <- function(x, ys) {
  sqrt((x - ys) ^ 2)
}
```


```{cpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector pdistC2(double x, NumericVector ys) {
  return sqrt(pow((x - ys), 2));
}
```

::: notes
Won't be faster than our C loop version but is easier to read
:::

---


## Other important C++ syntax

- In C++, vector indices start at 0!
- Use = for assignment
- C++ provides operators that modify in-place: total += x[i] is equivalent to total = total + x[i]. Similar in-place operators are -=, *=, and /=.
- Weird things can happen if you have missing values in an R object and pass them to C++


I am glossing over a **ton** of stuff. There is good documentation if you need to implement this yourself.

---

## Missing values


\fontsize{7pt}{8pt}\selectfont
R's `NA` does not map cleanly to C++ — each type has its own sentinel:

```{cpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
List scalar_missings() {
  int    int_s  = NA_INTEGER;
  String chr_s  = NA_STRING;
  bool   lgl_s  = NA_LOGICAL;
  double dbl_s  = NA_REAL;
  return List::create(int_s, chr_s, lgl_s, dbl_s);
}
```

For vectors, use `NumericVector::is_na()` to check:

```{cpp}
// [[Rcpp::export]]
LogicalVector is_naC(NumericVector x) {
  int n = x.size();
  LogicalVector out(n);
  for (int i = 0; i < n; ++i) {
    out[i] = NumericVector::is_na(x[i]);
  }
  return out;
}
```

::: notes
Passing an NA through without checking can silently produce wrong answers in C++ — no error thrown. Always think about whether your input might contain NAs.
:::

---

## Working with Lists and DataFrames


\fontsize{7pt}{8pt}\selectfont
Rcpp provides `List` and `DataFrame` types that map to R's `list` and `data.frame`:

```{cpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
List make_list(NumericVector x) {
  double mn = mean(x);
  double sd = Rcpp::sd(x);
  return List::create(Named("mean") = mn,
                      Named("sd")   = sd,
                      Named("n")    = x.size());
}
```

Reading from an existing R list passed in:

```{cpp}
// [[Rcpp::export]]
double list_sum(List lst) {
  NumericVector x = lst["x"];
  NumericVector y = lst["y"];
  return sum(x) + sum(y);
}
```

::: notes
DataFrame works similarly — DataFrame::create(Named("col") = vec). You can also accept a DataFrame argument and extract columns with df["colname"].
:::

---


## Debugging Rcpp code


\fontsize{9pt}{10pt}\selectfont
Rcpp is hard to debug. A few strategies:

**1. Write a pure-R version first** and test edge cases before porting to C++.

**2. Use `Rcpp::Rcout` instead of `std::cout`** to print from C++:

```{cpp}
// [[Rcpp::export]]
void debug_example(NumericVector x) {
  for (int i = 0; i < x.size(); ++i) {
    Rcpp::Rcout << "x[" << i << "] = " << x[i] << "\n";
  }
}
```

**3. Use `Rcpp::stop()` to throw errors** with informative messages:

```{cpp}
if (x.size() == 0) Rcpp::stop("x must be non-empty");
```

**4. Check types carefully** — passing the wrong R type (e.g., integer vs double) is a common silent bug.



::: notes
The most common beginner mistake is off-by-one errors (C++ is 0-indexed, R is 1-indexed) and type mismatches. Compile early and often.
:::

---

## Using Rcpp in a package


\fontsize{10pt}{11pt}\selectfont
To add `Rcpp` to an existing package, put your C++ files in the `src/` directory and:

- In `DESCRIPTION` add:
```
LinkingTo: Rcpp
Imports: Rcpp
```

- In `NAMESPACE` add (or let `roxygen2` handle it):
```
importFrom(Rcpp, evalCpp)
useDynLib(yourpkg, .registration = TRUE)
```

- The easiest way:
```{r}
usethis::use_rcpp()
```

This edits `DESCRIPTION`, `NAMESPACE`, creates `src/`, and adds a placeholder `.cpp` file.

::: notes
For more details see the Rcpp package vignette, `vignette("Rcpp-package")`.
:::


---

## Final thoughts on converting R code to C++

- Test everything!!
  - Check that the output of your code in C++ matches the output of your code in R
- `RcppArmadillo` is a package with nice linear algebra syntax
  - Makes it easier to implement C++ code that uses a lot of matrix operations

- ChatGPT probably makes this a lot easier... but still test your code

---

## My package development workflow

- load `devtools()`
- `load_all()`
- `document()`
- `check()`
- `install()`

---



## References

- Wickham H, Bryan J. [*R Packages* (2e).](https://r-pkgs.org)
- Wickham H. [*Advanced R* (2e), Chapter on R's package system.](https://adv-r.hadley.nz)
- [`usethis` documentation](https://usethis.r-lib.org)
- [`devtools` documentation](https://devtools.r-lib.org)
- Eddelbuettel D. [*Seamless R and C++ Integration with Rcpp*.](https://www.rcpp.org/book/) Springer.
- [Rcpp vignettes](https://cran.r-project.org/web/packages/Rcpp/vignettes/)
